Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-26 2024/OPTIMA/OPTIMA_2024_MCO.tex
fix some misprints

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by by the Research and Education Mathematics Centre ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriterial problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \Letter\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

%
\begin{abstract}
To solve a multicriterial optimization problem, it is necessary to find a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for solving black-box multiobjective problems with multiextremal criteria. The developed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriterial optimization. The effectiveness of different versions of the developed algorithm is estimated on the base of  a representative computational experiment.

\keywords{Multicriterial Problems \and Global Optimization \and Machine Learning.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:1}

@@

In our study we deal with methods for solving the most complex class of multicriterial problems where criteria are considered as ``black-box'' functions and are multiextremal \cite{Miettinen1999,Ehrgott2005,Pardalos2017,Strongin2000,Sergeyev2013}.

@@

In the paper \cite{ML_MCO_2023} we proposed an approach to solving multicriterial optimization problems combining the classic information-statistical global search algorithm with machine learning methods. In the framework of the approach a problem of multicriterial optimization was transformed to a series of scalar optimization subproblems. In the course of solving scalar subproblems, on the base of machine learning methods a hyperplane was built and gradually refined which separated points of effective and ineffective solutions in the criteria space. Additional information connected with the separating hyperplane was embedded in the computational scheme of the optimization algorithm and, as a result, the number of trials (criteria evaluations) in domains with ineffective solutions was significantly decreased. The experimental results showed that in comparison with a number of heuristic algorithms, the search for effective solutions was several times faster. At the same time, this approach can be improved because utilizing only the separating plane limits the class of machine learning techniques which can be applied for enhancing the global search algorithm.

@@

The paper has the following structure. Section \ref{sec:2} describes the mathematical statement of the problem to be studied. Section \ref{sec:3} is devoted to a brief description of the basic algorithms used for solving problems of multicriterial optimization. Section \ref{sec:4} contains information on the techniques applied for reducing the number of trials during the optimization process. Subsection \ref{ssec:41} gives a short description of the method considered in \cite{ML_MCO_2023}. A new approach to using probabilities of trial points belonging to different efficiency classes is presented in subsection \ref{ssec:42}. Section \ref{sec:5} contains results of computational experiments demonstrating the effectiveness of the proposed approach.

@@

The multicriterial optimization (MCO) problem can be formulated as the minimization problem of the vector-function. Let us introduce the vector-function (vector criterion)

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \in R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, where $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.

The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.

@@

As it was mentioned above, the partial criteria contradictoriness leads to the necessity to introduce the set of non-dominated points   called Pareto set as the general solution of the multicriterial problem \cite{Miettinen1999,Ehrgott2005,Pardalos2017}. Hereinafter, any point of the Pareto set will be called a partial solution or just an effective point.

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriteria problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by by the Research and Education Mathematics Centre ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriteria problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \Letter\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

%
\begin{abstract}
To solve a multicriteria optimization problem, it is necessary to find a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for solving black-box multiobjective problems with multiextremal criteria. The developed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriteria optimization. The effectiveness of different versions of the developed algorithm is estimated on the base of  a representative computational experiment.

\keywords{Multicriteria Problems \and Global Optimization \and Machine Learning.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:1}

@@

In our study we deal with methods for solving the most complex class of multicriteria problems where criteria are considered as ``black-box'' functions and are multiextremal \cite{Miettinen1999,Ehrgott2005,Pardalos2017,Strongin2000,Sergeyev2013}.

@@

In the paper \cite{ML_MCO_2023} we proposed an approach to solving multicriteria optimization problems combining the classic information-statistical global search algorithm with machine learning methods. In the framework of the approach a problem of multicriteria optimization was transformed to a series of scalar optimization subproblems. In the course of solving scalar subproblems, on the base of machine learning methods a hyperplane was built and gradually refined which separated points of effective and ineffective solutions in the criteria space. Additional information connected with the separating hyperplane was embedded in the computational scheme of the optimization algorithm and, as a result, the number of trials (criteria evaluations) in domains with ineffective solutions was significantly decreased. The experimental results showed that in comparison with a number of heuristic algorithms, the search for effective solutions was several times faster. At the same time, this approach can be improved because utilizing only the separating plane limits the class of machine learning techniques which can be applied for enhancing the global search algorithm.

@@

The paper has the following structure. Section \ref{sec:2} describes the mathematical statement of the problem to be studied. Section \ref{sec:3} is devoted to a brief description of the basic algorithms used for solving problems of multicriteria optimization. Section \ref{sec:4} contains information on the techniques applied for reducing the number of trials during the optimization process. Subsection \ref{ssec:41} gives a short description of the method considered in \cite{ML_MCO_2023}. A new approach to using probabilities of trial points belonging to different efficiency classes is presented in subsection \ref{ssec:42}. Section \ref{sec:5} contains results of computational experiments demonstrating the effectiveness of the proposed approach.

@@

The multicriteria optimization (MCO) problem can be formulated as the minimization problem of the vector-function. Let us introduce the vector-function (vector criterion)

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \in R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, and $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.
The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.

@@

As it was mentioned above, the partial criteria contradictoriness leads to the necessity to introduce the set of non-dominated points called Pareto set as the general solution of the multicriteria problem \cite{Miettinen1999,Ehrgott2005,Pardalos2017}. Hereinafter, any point of the Pareto set will be called a \textit{partial solution} or just an \textit{efficient point}.
